### 2023-10-14 chapters/paper_benchmarking.tex
Fix many spelling and language errors

To advance the state of LVMs for speech, this paper (i) develops a benchmark for LVMs based on model likelihood, (ii) introduces a hierarchical LVM architecture without autoregressive decoder, (iii) compares LVMs to deterministic counterparts including WaveNet, and (iv) qualitatively and quantitatively evaluates the latent variables learned by different LVMs based on their usefulness for phoneme recognition. We find that:
\begin{itemize}[leftmargin=8mm]
    \setlength\itemsep{-0em}
    \item [(I)] State-of-the-art LVMs achieve likelihoods that are inferior to WaveNet at high temporal resolution but are superior at lower resolutions. Interestingly, we find that a standard LSTM \parencite{hochreiter_long_1997} almost matches the likelihood of WaveNet.
    \item [(II)] LVMs with powerful autoregressive decoders achieve better likelihoods than the non"=autoregressive LVM.
    \item[(III)] The expressiveness of LVMs for speech increases with a deeper hierarchy of stochastic latent variables, similar to conclusions within image modeling. % \parencite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
    \item[(IV)] LVMs learn rich representations that are as good or better than Mel spectrograms for phoneme recognition also when using only 10 minutes of labeled data.
\end{itemize}
At a high level, this benchmark brings order to LVM model comparisons for speech and also provides useful reference implementations of the models
\footnote{\href{https://github.com/JakobHavtorn/benchmarking-lvms}{\url{github.com/JakobHavtorn/benchmarking-lvms}}}.
% \footnote{\fontsize{8.5}{0}\selectfont\texttt{ \url{github.com/anonymous/available-later}}}.
Before presenting the results, we provide a brief survey of existing LVMs for speech in a coherent notation.

@@

                \edge[]{d_t}{z_t};
                \edge[bend right]{d_t}{x_t};
                \edge[]{z_tm1}{d_t};
                \edge[]{d_tm1}{d_t};
                \edge[]{z_t}{x_t};
            \end{tikzpicture}
        }
        \end{multicols}
    \caption[Generative models for an LSTM, the VRNN, SRNN, STCN, and CW-VAE models.]{
        Generative models for a single time step of a deterministic autoregressive LSTM, the VRNN and SRNN as well as the STCN and CW-VAE both with a single layer of latent variables. 
        Red arrows indicate purely deterministic paths from the output $\xb_t$ to previous input $\xb_{<t}$ without passing a stochastic node. 
        The models differ in their use of latent variables and dependencies especially autoregressive dependencies and skip connections. 
        % In an LSTM, information flows deterministically from previous observed variables to the next. 
        % The VRNN and SRNN use stochastic latent variables but also include deterministic skip connections from previous observed variables. 
        % The STCN is also autoregressive in $\xb$ but does not use deterministic skip connections from $\xb_{t-1}$ to $\xb_t$. 
        % The CW-VAE is not autoregressive on the observed variable which forces information to flow through the latent variables. 
        We provide more elaborate graphical illustrations including inference models in \cref{fig: clockwork vae inference generative models 2 layered} and appendix~\cref{app: additional graphical models}.
        }
    \label{fig: autoregressive vrnn srnn cwvae 1L graphs}
\end{figure*}

@@

First, $\xb_{1:T}$ is encoded to a temporal stochastic latent representation $\zb_{1:T} = \left( \zb_1, \zb_2, \dots, \zb_{T} \right)$ with $\zb_t\in \mathbb{R}^{D_z}$. This representation is then used to reconstruct the original input $\xb_{1:T}$.
The latent variable is assumed to follow some prior distribution $p(\zb_t|\cdot)$ where the dot indicates that it may depend on latent and observed variables at previous time steps, $\zb_{<t}$ and $\xb_{<t}$ where $\zb_{<t} := \left( \zb_1, \zb_2, \dots, \zb_{t-1} \right)$.

@@

            % \node[above=of s_2_1, yshift=-1.cm, xshift=2cm] (phi) {$p\pa{\xb_{1:T},\zb_{1:T}^{(1:L)}}$};
            \node[above=of s_2_1, yshift=-0.5cm, xshift=2.5cm] (phi) {CW-VAE generative};
        \end{tikzpicture}
    % \hfill
    }
\caption[Generative and inference models for a two-layered CW-VAE.]{
Inference (left) and generative (right) models for the CW-VAE with a hierarchy of $L=2$ latent variables, $s_1=1$ and $s_2=2$. 
The models are unrolled over four consecutive time steps but note that the graph continues towards $t=0$ and $t=T$. 
Blue arrows indicate parameter sharing between inference and generative models. 
We omit the deterministic variable of \cref{fig: autoregressive vrnn srnn cwvae 1L graphs} for clarity.
}
\label{fig: clockwork vae inference generative models 2 layered}
\end{figure*}

@@

To simplify temporal notation, we define the timesteps at which a layer updates its latent state as $\mathcal{T}_l := \{t \in \bra{1,T} \,|\, (t-1)\,\text{mod}\, s_l = 0\}$. We then define the set of layers that update at a given timestep as $\mathcal{J}_t := \{ l \,|\, t\in\mathcal{T}_l \}$. The joint distribution can then be written as,
% The timesteps at which a layer updates its latent state are given by $\mathcal{T}_l := \{t \in \cbra{1,\dots,T} ~|~ t\mod s_l = 1\}$.
% To simplify temporal notation, we equivalently define this by copying references to the unique states over time, $\zb^{(l)}_t := \zb^{(l)}_{u}$ with $u=\max_\tau \cbra{\tau \in \mathcal{T}_l ~|~ \tau \leq t}$.
% The joint distribution factorizes over time and the hierarchy.
\begin{equation}
    % p\pa{\xb_{1:T},\zb_{1:T}} = \prod_{t} p\pa{\xb_t | \zb_t^{(1)}} \prod_{l=1}^L\prod_{t\in\mathcal{T}_l} p\pa{\zb_t^{(l)} | \zb_{t-1}^{(l)}, \zb_t^{(l+1)}} \enspace .
    p(\xb_{1:T},\zb_{1:T}^{(1:L)}) = 
    \prod_{t=1}^{T} p(\xb_t | \zb_t^{(1)})
    \prod\limits_{l\in\mathcal{J}_t} p(\zb_t^{(l)} | \zb_{t-1}^{(l)}, \zb_t^{(l+1)}) \enspace . \nonumber
    \label{eq: cwvae joint}
\end{equation}
The inference model is conditioned on a window of the observed variable $\xb_{t:t+s_l}$ depending on the layer's stride $s_l$.
\begin{equation}
    q\pa{\zb_{1:T} | \xb_{1:T}} = \prod_{t=1}^T \prod\limits_{l\in\mathcal{J}_t} q\pa{\zb_t^{(l)} | \zb_{t-1}^{(l)}, \zb_t^{(l+1)}, \xb_{t:t+s_l}} \enspace . \nonumber
    \label{eq: cwvae inference}
\end{equation}
The dependency on $\xb_{t:t+s_l}$ is encoded via a convolutional ladder network similar to the STCN with $\db_t^{(l)} = e(\xb_{t:t+s_l})$. We define $\xb_{\scriptscriptstyle\mathcal{S}_t^l}:=\xb_{t:t+s_l}$ for compactness. The latent state transitions are densely connected and the decoder is also a convolutional network. 
% \begin{align}
%     q\left(\zb_t^{(l)}|\xb_{\scriptscriptstyle\mathcal{S}_t^l},\zb_{t-1}^{(l)},\zb_t^{(l+1)}\right) &= \mathcal{N}\pa{\alphab_{q}^{l}\pa{\zb_{t-1}^{(l)},\zb_{t}^{(l+1)},\db_t^{(l)}}} \nonumber \\ % \label{eq: cwvae parameterization inference} \\
%     p\left(\zb_t^{(l)}|\zb_{t-1}^{(l)},\zb_{t}^{(l+1)}\right) &= \mathcal{N}\pa{\alphab_{p}^{l}\pa{\zb_{t-1}^{(l)},\zb_t^{(l+1)}}} \nonumber \\% \label{eq: cwvae parameterization prior} \\
%     p\left(\xb_t|\zb^{(1)}_{t-s_l/2:t+s_l/2}\right) &= \mathcal{D}\pa{\betab\pa{\zb^{(1)}_{t-s_l/2:t+s_l/2}}} \nonumber % \enspace . \label{eq: cwvae parameterization observation}
% \end{align}
\begin{gather}
    p\left(\zb_t^{(l)}|\zb_{t-1}^{(l)},\zb_{t}^{(l+1)}\right) = \mathcal{N}\pa{\alphab_{p}^{l}\pa{\zb_{t-1}^{(l)},\zb_t^{(l+1)}}} \, , \\ \nonumber
    p\left(\xb_t|\zb^{(1)}_{t-s_l/2:t+s_l/2}\right) = \mathcal{D}\pa{\betab\pa{\zb^{(1)}_{t-s_l/2:t+s_l/2}}} \, , \\
    q\left(\zb_t^{(l)}|\xb_{\scriptscriptstyle\mathcal{S}_t^l},\zb_{t-1}^{(l)},\zb_t^{(l+1)}\right) = \mathcal{N}\pa{\alphab_{q}^{l}\pa{\zb_{t-1}^{(l)},\zb_{t}^{(l+1)},\db_t^{(l)}}} \nonumber
\end{gather}

@@

\section{Speech modeling benchmark}\label{sec:glvm_benchmark}
\paragraph{Data}
We train models on TIMIT \parencite{garofolo_timit_1993}, LibriSpeech \parencite{panayotov_librispeech_2015} and LibriLight \parencite{kahn_libri-light_2020}. For TIMIT, we randomly sample 5\% of the training split for validation. 
We represent the audio as $\mu$-law encoded PCM standardized to values in $\bra{-1, 1}$ with discretization gap of ${2^{-b+1}}$. We use the original bit depth of $\SI{16}{bits}$ and sample rate of $\SI{16000}{Hz}$. We use this representation both as the input and the reconstruction target. We provide more details on the datasets in appendix~\cref{app: dataset details} and additional results on linear PCM in appendix~\cref{app: additional likelihood results appendix}.
%
\paragraph{Likelihood}
We report likelihoods in units of bits per frame (bpf) as this yields a more interpretable and comparable likelihood than total likelihood in nats. It also has direct connections with information theory and compression \parencite{shannon_mathematical_1948, townsend_practical_2019}. In units of bits per frame, lower is better. For LVMs, we report the one-sample ELBO. The likelihoods can be seen in \cref{tab: likelihoods timit,tab: likelihoods librispeech}. We describe how to convert likelihood to bpf in appendix~\cref{app: likelihood in bits per frame}.
%
\paragraph{Models}
Architecture and training details are sketched below, while the full details are in appendices~\cref{app: model architectures} and \cref{app: training details} along with additional results for some alternative model configurations in appendix~\cref{app: additional likelihood results appendix}.
We select model configurations that can be trained on GPUs with a maximum of 12GB of RAM and train all models until convergence on the validation set.
We use a DMoL with 10 components for the output distribution of all models and model all datasets at their full bit depth of $\SI{16}{bits}$. We train and evaluate models on stacked waveforms similar to previous work \parencite{chung_recurrent_2015, fraccaro_sequential_2016, aksan_stcn_2019} with stack sizes of $s=1$, $s=64$ and $s=256$. Hence, every model input $\xb_t$ is composed of $\tilde{x}_{t:t+s}$ where $\tilde{x}$ are waveform frames. We provide additional results with a Gaussian output distribution in appendix~\cref{app: additional likelihood results appendix}.

@@

    \caption[Model likelihoods and phoneme error rate for TIMIT.]{We report model likelihoods $\mathcal{L}$ for TIMIT represented as a 16bit $\mu$-law encoded PCM for different stochastic latent variable models and deterministic autoregressive baselines (left) and phoneme error rate (PER) of different representations for phoneme recognition on TIMIT (right).}

@@

        3.7h   &  Spectrogram     & Mel           &   24.1   \\  % 2wjqhwf7  0.24149049930081432
        3.7h   &  WaveNet         & $\hb^{(15)}$  &   27.7   \\  % 36ccyxf4  0.27743686764826847
        3.7h   &  LSTM            & $\hb$         &   23.0   \\  % 3bh4vgrx  0.2301719174138357
        3.7h   &  VRNN            & $\zb$         &   23.2   \\  % 3stwdgas  0.23229415151764415
        3.7h   &  SRNN            & $\zb$         &   26.0   \\  % 2bkvv5i0  0.2596528748868964
        3.7h   &  CW-VAE          & $\zb^{(1)}$   &   36.4   \\  % 3m4f3f7z
        3.7h   &  STCN            & $\zb^{(2)}$   &   \textbf{21.9}   \\  % rkukk5bn  0.2189849469441474

@@

        1.0h   &  Spectrogram     & Mel           &   30.8   \\  % 3kmfpljj  0.3083655507115242
        1.0h   &  WaveNet         & $\hb^{(15)}$  &   34.7   \\  % 36ccyxf4  0.34743769021962656
        1.0h   &  LSTM            & $\hb$         &   30.1   \\  % 3bh4vgrx  0.30137369416796905
        1.0h   &  VRNN            & $\zb$         &   30.4   \\  % 3stwdgas  0.31284033889939955
        1.0h   &  SRNN            & $\zb$         &   31.7   \\  % 2bkvv5i0  0.3167064242823065
        1.0h   &  CW-VAE          & $\zb^{(1)}$   &   40.0   \\  % 3m4f3f7z
        1.0h   &  STCN            & $\zb^{(2)}$   &   \textbf{26.7}   \\  % rkukk5bn  0.2671547256724521

@@

        10m    &  Waveform        & stacked       &   85.6   \\  %           0.8564448465904417
        10m    &  Spectrogram     & Mel           &   47.1   \\  % 1sep2uq8  0.4706753310849716
        10m    &  WaveNet         & $\hb^{(15)}$  &   52.8   \\  % 36ccyxf4  0.5276137204902526
        10m    &  LSTM            & $\hb$         &   46.1   \\  % 3bh4vgrx  0.46073866907954264
        10m    &  VRNN            & $\zb$         &   44.6   \\  % 3stwdgas  0.4601299662745743
        10m    &  SRNN            & $\zb$         &   47.3   \\  % 2bkvv5i0  0.47317594801349017
        10m    &  CW-VAE          & $\zb^{(1)}$   &   54.9   \\  % 3m4f3f7z
        10m    &  STCN            & $\zb^{(2)}$   &   \textbf{42.7}   \\  % rkukk5bn  0,4270625976803488

@@

        1 & Wavenet             & $D_c=96$       & \bfseries 10.96/10.89 & \bfseries 10.85/10.76 & \bfseries 11.12/11.01 & \bfseries 11.05/10.85 \\

@@

        64 & Wavenet            & $D_c=96$       & 13.61/13.24 & 13.58/13.21 & 13.61/13.22 & 13.60/13.21 \\  % 2ghkybtd/3iieasdt

@@

At $s=64$, Wavenet and the LSTM yield significantly worse likelihoods than all LVMs separated by $\sim\SI{1}{bit}$.
The CW-VAE outperforms the VRNN and SRNN when configured with a hierarchy of latent variables. 
With a single layer of latent variables, the CW-VAE is inferior to both SRNN and VRNN but notably still better than the LSTM. 
These observations carry to $s=256$, where a multilayered CW-VAE outperforms the LSTM, VRNN and SRNN. 
The STCN yields the best results at both $s=64$ and $s=256$. 
For strides $s>1$, previous work has attributed the inferior performance of autoregressive models without latent variables, such as WaveNet and the LSTM, to the ability of LVMs to model intra-step correlations \parencite{lai_re-examination_2019}.

@@

\section{Phoneme recognition}\label{sec: phoneme recognition}
Although the likelihood is a practical metric for model comparison and selection, a high likelihood does not guarantee that a model has learned useful representations \parencite{huszar_is_2017}. 
For speech data, we would expect models to learn features related to phonetics which would make them useful for tasks such as automatic speech recognition (ASR). 
The Mel spectrogram is a well-established representation of audio designed for speech recognition and is predefined rather than learned. % \parencite{??}. Basic Correlates of the Auditory Stimulus, https://psycnet.apa.org/record/1951-07758-017, modern mel scale https://ieeexplore.ieee.org/stamp/stamp.jsp?arnumber=1170013&casa_token=UIyQaj3tBRAAAAAA:GtttZedngBkaxVumjnXctc5zS50fptexPMiWhr-7gDrmIVq3JqJPnGCWYedoymb0X5JTGElmHf8&tag=1,
To assess the usefulness of the representations learned by the benchmarked models, we compare them to the highly useful Mel spectrogram on the task of phoneme recognition. 
Phonemes are fundamental units of speech that relate to how parts of words are pronounced (see also appendix~\cref{app: timit phoneme distributions}).
% rather than characters or words themselves (see also appendix~\cref{apsp: timit phoneme distributions}).

\paragraph{Quantitative}
We train an ASR model to recognize phonemes and compare its performance when using input representations obtained from different unsupervised models. 
For WaveNet and the LSTM, we use the hidden state as the representation. For all LVMs, we use the latent variable. For the hierarchical LVMs and WaveNet, we run the experiment using each possible representation and report only the best one here. 
We compare the learned representations to a log Mel spectrogram with 80 filterbanks, hop length 64 and window size 128. 
We also compared to using raw PCM in vectors of 64 elements standardized to $[-1,1]$ but found that the ASR did not reliably converge at all which highlights the importance of input representation. 
The ASR model is a three-layered bidirectional LSTM with 256 hidden units. It is trained with the connectionist temporal classification (CTC) loss \parencite{graves_connectionist_2006} which lets it jointly learn to align and classify without using label alignments. 
We pre"=train all unsupervised models at $s=64$ on the full TIMIT training dataset excluding the validation data (3.7h) as in \cref{tab: likelihoods timit} (left). 
We then train the ASR model on all 3.7h as well as 1h and 10m subsets. We report results on the test set in terms of phoneme error rates (PER) in \cref{tab: phoneme recognition (PER)} (right).

>>>>

To advance the state of LVMs for speech, this paper (i) develops a benchmark for LVMs based on model likelihood, (ii) introduces a hierarchical LVM architecture without autoregressive decoder, (iii) compares LVMs to deterministic counterparts including WaveNet, and (iv) qualitatively and quantitatively evaluates the latent variables learned by different LVMs based on their usefulness for phoneme recognition. We find that:
\begin{itemize}[leftmargin=8mm]
    \setlength\itemsep{-0em}
    \item [(I)] State-of-the-art LVMs achieve likelihoods that are inferior to WaveNet at high temporal resolution but are superior at lower resolutions. Interestingly, we find that a standard LSTM \parencite{hochreiter_long_1997} almost matches the likelihood of WaveNet.
    \item [(II)] LVMs with powerful autoregressive decoders achieve better likelihoods than the non"=autoregressive LVM.
    \item[(III)] The expressiveness of LVMs for speech increases with a deeper hierarchy of stochastic latent variables, similar to conclusions within image modeling. % \parencite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
    \item[(IV)] LVMs learn rich representations that are as good or better than Mel spectrograms for phoneme recognition also when using only 10 minutes of labeled data.
\end{itemize}
At a high level, this benchmark brings order to LVM model comparisons for speech and also provides useful reference implementations of the models\footnote{\href{https://github.com/JakobHavtorn/benchmarking-lvms}{\url{github.com/JakobHavtorn/benchmarking-lvms}}}.
% \footnote{\fontsize{8.5}{0}\selectfont\texttt{ \url{github.com/anonymous/available-later}}}.
Before presenting the results, we provide a brief survey of existing LVMs for speech in a coherent notation.

@@

                \edge[]{d_t}{z_t};
                \edge[bend right]{d_t}{x_t};
                \edge[]{z_tm1}{d_t};
                \edge[]{d_tm1}{d_t};
                \edge[]{z_t}{x_t};
            \end{tikzpicture}
        }
        \end{multicols}
    \caption[Generative models for an LSTM, the VRNN, SRNN, STCN, and CW-VAE models.]{
        Generative models for a single timestep of a deterministic autoregressive LSTM, the VRNN and SRNN as well as the STCN and CW-VAE both with a single layer of latent variables. 
        Red arrows indicate purely deterministic paths from the output $\xb_t$ to previous input $\xb_{<t}$ without passing a stochastic node. 
        The models differ in their use of latent variables and dependencies especially autoregressive dependencies and skip connections. 
        % In an LSTM, information flows deterministically from previous observed variables to the next. 
        % The VRNN and SRNN use stochastic latent variables but also include deterministic skip connections from previous observed variables. 
        % The STCN is also autoregressive in $\xb$ but does not use deterministic skip connections from $\xb_{t-1}$ to $\xb_t$. 
        % The CW-VAE is not autoregressive on the observed variable which forces information to flow through the latent variables. 
        We provide more elaborate graphical illustrations including inference models in \cref{fig: clockwork vae inference generative models 2 layered} and appendix~\cref{app: additional graphical models}.
        }
    \label{fig: autoregressive vrnn srnn cwvae 1L graphs}
\end{figure*}

@@

First, $\xb_{1:T}$ is encoded to a temporal stochastic latent representation $\zb_{1:T} = \left( \zb_1, \zb_2, \dots, \zb_{T} \right)$ with $\zb_t\in \mathbb{R}^{D_z}$. This representation is then used to reconstruct the original input $\xb_{1:T}$.
The latent variable is assumed to follow some prior distribution $p(\zb_t|\cdot)$ where the dot indicates that it may depend on latent and observed variables at previous timesteps, $\zb_{<t}$ and $\xb_{<t}$ where $\zb_{<t} := \left( \zb_1, \zb_2, \dots, \zb_{t-1} \right)$.

@@

            % \node[above=of s_2_1, yshift=-1.cm, xshift=2cm] (phi) {$p\pa{\xb_{1:T},\zb_{1:T}^{(1:L)}}$};
            \node[above=of s_2_1, yshift=-0.5cm, xshift=2.5cm] (phi) {CW-VAE generative};
        \end{tikzpicture}
    % \hfill
    }
\caption[Generative and inference models for a two-layered CW-VAE.]{
Inference (left) and generative (right) models for the CW-VAE with a hierarchy of $L=2$ latent variables, $s_1=1$ and $s_2=2$. 
The models are unrolled over four consecutive timesteps but note that the graph continues towards $t=0$ and $t=T$. 
Blue arrows indicate parameter sharing between inference and generative models. 
We omit the deterministic variable of \cref{fig: autoregressive vrnn srnn cwvae 1L graphs} for clarity.
}
\label{fig: clockwork vae inference generative models 2 layered}
\end{figure*}

@@

To simplify temporal notation, we define the timesteps at which a layer updates its latent state as $\mathcal{T}_l := \{t \in \bra{1,T} \,|\, (t-1)\,\text{mod}\, s_l = 0\}$. We then define the set of layers that update at a given timestep as $\mathcal{J}_t := \{ l \,|\, t\in\mathcal{T}_l \}$. The joint distribution can then be written as,
% The timesteps at which a layer updates its latent state are given by $\mathcal{T}_l := \{t \in \cbra{1,\dots,T} ~|~ t\mod s_l = 1\}$.
% To simplify temporal notation, we equivalently define this by copying references to the unique states over time, $\zb^{(l)}_t := \zb^{(l)}_{u}$ with $u=\max_\tau \cbra{\tau \in \mathcal{T}_l ~|~ \tau \leq t}$.
% The joint distribution factorizes over time and the hierarchy.
\begin{equation}
    % p\pa{\xb_{1:T},\zb_{1:T}} = \prod_{t} p\pa{\xb_t | \zb_t^{(1)}} \prod_{l=1}^L\prod_{t\in\mathcal{T}_l} p\pa{\zb_t^{(l)} | \zb_{t-1}^{(l)}, \zb_t^{(l+1)}} \enspace .
    p(\xb_{1:T},\zb_{1:T}^{(1:L)}) = 
    \prod_{t=1}^{T} p(\xb_t | \zb_t^{(1)})
    \prod\limits_{l\in\mathcal{J}_t} p(\zb_t^{(l)} | \zb_{t-1}^{(l)}, \zb_t^{(l+1)}) \enspace . \nonumber
    \label{eq: cwvae joint}
\end{equation}
The inference model is conditioned on a window of the observed variable $\xb_{t:t+s_l}$ depending on the layer's stride $s_l$.
\begin{equation}
    q\pa{\zb_{1:T} | \xb_{1:T}} = \prod_{t=1}^T \prod\limits_{l\in\mathcal{J}_t} q\pa{\zb_t^{(l)} | \zb_{t-1}^{(l)}, \zb_t^{(l+1)}, \xb_{t:t+s_l}} \enspace . \nonumber
    \label{eq: cwvae inference}
\end{equation}
The dependency on $\xb_{t:t+s_l}$ is encoded via a convolutional ladder network similar to the STCN with $\db_t^{(l)} = e(\xb_{t:t+s_l})$. We define $\xb_{\scriptscriptstyle\mathcal{S}_t^l}:=\xb_{t:t+s_l}$ for compactness. The latent state transitions are densely connected, and the decoder is also a convolutional network. 
% \begin{align}
%     q\left(\zb_t^{(l)}|\xb_{\scriptscriptstyle\mathcal{S}_t^l},\zb_{t-1}^{(l)},\zb_t^{(l+1)}\right) &= \mathcal{N}\pa{\alphab_{q}^{l}\pa{\zb_{t-1}^{(l)},\zb_{t}^{(l+1)},\db_t^{(l)}}} \nonumber \\ % \label{eq: cwvae parameterization inference} \\
%     p\left(\zb_t^{(l)}|\zb_{t-1}^{(l)},\zb_{t}^{(l+1)}\right) &= \mathcal{N}\pa{\alphab_{p}^{l}\pa{\zb_{t-1}^{(l)},\zb_t^{(l+1)}}} \nonumber \\% \label{eq: cwvae parameterization prior} \\
%     p\left(\xb_t|\zb^{(1)}_{t-s_l/2:t+s_l/2}\right) &= \mathcal{D}\pa{\betab\pa{\zb^{(1)}_{t-s_l/2:t+s_l/2}}} \nonumber % \enspace . \label{eq: cwvae parameterization observation}
% \end{align}
\begin{gather}
    p\left(\zb_t^{(l)}|\zb_{t-1}^{(l)},\zb_{t}^{(l+1)}\right) = \mathcal{N}\pa{\alphab_{p}^{l}\pa{\zb_{t-1}^{(l)},\zb_t^{(l+1)}}} \, , \\ \nonumber
    p\left(\xb_t|\zb^{(1)}_{t-s_l/2:t+s_l/2}\right) = \mathcal{D}\pa{\betab\pa{\zb^{(1)}_{t-s_l/2:t+s_l/2}}} \, , \\
    q\left(\zb_t^{(l)}|\xb_{\scriptscriptstyle\mathcal{S}_t^l},\zb_{t-1}^{(l)},\zb_t^{(l+1)}\right) = \mathcal{N}\pa{\alphab_{q}^{l}\pa{\zb_{t-1}^{(l)},\zb_{t}^{(l+1)},\db_t^{(l)}}} \nonumber
\end{gather}

@@

\section{Speech modeling benchmark}\label{sec:glvm_benchmark}
\paragraph{Data}
We train models on TIMIT \parencite{garofolo_timit_1993}, LibriSpeech \parencite{panayotov_librispeech_2015} and LibriLight \parencite{kahn_libri-light_2020}. For TIMIT, we randomly sample 5\% of the training split for validation. 
We represent the audio as $\mu$-law encoded PCM standardized to values in $\bra{-1, 1}$ with discretization gap of ${2^{-b+1}}$. We use the original bit depth of $\SI{16}{bits}$ and sample rate of $\SI{16000}{Hz}$. We use this representation both as the input and the reconstruction target. We provide more details on the datasets in appendix~\cref{app: dataset details} and additional results on linear PCM in appendix~\cref{app: additional likelihood results appendix}.
%
\paragraph{Likelihood}
We report likelihoods in units of bits per frame (bpf) as this yields a more interpretable and comparable likelihood than total likelihood in nats. It also has direct connections with information theory and compression \parencite{shannon_mathematical_1948, townsend_practical_2019}. In units of bits per frame, lower is better. For LVMs, we report the one-sample ELBO. The likelihoods can be seen in \cref{tab: likelihoods timit,tab: likelihoods librispeech}. We describe how to convert likelihood to bpf in appendix~\cref{app: likelihood in bits per frame}.
%
\paragraph{Models}
Architecture and training details are sketched below, while the full details are in appendices~\cref{app: model architectures} and \cref{app: training details} along with additional results for some alternative model configurations in appendix~\cref{app: additional likelihood results appendix}.
We select model configurations that can be trained on GPUs with a maximum of \SI{12}{GB} of RAM and train all models until convergence on the validation set.
We use a DMoL with 10 components for the output distribution of all models and model all datasets at their full bit depth of $\SI{16}{bits}$. We train and evaluate models on stacked waveforms similar to previous work \parencite{chung_recurrent_2015, fraccaro_sequential_2016, aksan_stcn_2019} with stack sizes of $s=1$, $s=64$ and $s=256$. Hence, every model input $\xb_t$ is composed of $\tilde{x}_{t:t+s}$ where $\tilde{x}$ are waveform frames. We provide additional results with a Gaussian output distribution in appendix~\cref{app: additional likelihood results appendix}.

@@

    \caption[Model likelihoods and phoneme error rate for TIMIT.]{ We report model likelihoods $\mathcal{L}$ for TIMIT represented as a 16bit $\mu$-law encoded PCM for different stochastic latent variable models and deterministic autoregressive baselines (left) and phoneme error rate (PER) of different representations for phoneme recognition on TIMIT (right).}

@@

        \SI{3.7}{h}   &  Spectrogram     & Mel           &   24.1   \\  % 2wjqhwf7  0.24149049930081432
        \SI{3.7}{h}   &  WaveNet         & $\hb^{(15)}$  &   27.7   \\  % 36ccyxf4  0.27743686764826847
        \SI{3.7}{h}   &  LSTM            & $\hb$         &   23.0   \\  % 3bh4vgrx  0.2301719174138357
        \SI{3.7}{h}   &  VRNN            & $\zb$         &   23.2   \\  % 3stwdgas  0.23229415151764415
        \SI{3.7}{h}   &  SRNN            & $\zb$         &   26.0   \\  % 2bkvv5i0  0.2596528748868964
        \SI{3.7}{h}   &  CW-VAE          & $\zb^{(1)}$   &   36.4   \\  % 3m4f3f7z
        \SI{3.7}{h}   &  STCN            & $\zb^{(2)}$   &   \textbf{21.9}   \\  % rkukk5bn  0.2189849469441474

@@

        \SI{1.0}{h}   &  Spectrogram     & Mel           &   30.8   \\  % 3kmfpljj  0.3083655507115242
        \SI{1.0}{h}   &  WaveNet         & $\hb^{(15)}$  &   34.7   \\  % 36ccyxf4  0.34743769021962656
        \SI{1.0}{h}   &  LSTM            & $\hb$         &   30.1   \\  % 3bh4vgrx  0.30137369416796905
        \SI{1.0}{h}   &  VRNN            & $\zb$         &   30.4   \\  % 3stwdgas  0.31284033889939955
        \SI{1.0}{h}   &  SRNN            & $\zb$         &   31.7   \\  % 2bkvv5i0  0.3167064242823065
        \SI{1.0}{h}   &  CW-VAE          & $\zb^{(1)}$   &   40.0   \\  % 3m4f3f7z
        \SI{1.0}{h}   &  STCN            & $\zb^{(2)}$   &   \textbf{26.7}   \\  % rkukk5bn  0.2671547256724521

@@

        \SI{10}{m}    &  Waveform        & stacked       &   85.6   \\  %           0.8564448465904417
        \SI{10}{m}    &  Spectrogram     & Mel           &   47.1   \\  % 1sep2uq8  0.4706753310849716
        \SI{10}{m}    &  WaveNet         & $\hb^{(15)}$  &   52.8   \\  % 36ccyxf4  0.5276137204902526
        \SI{10}{m}    &  LSTM            & $\hb$         &   46.1   \\  % 3bh4vgrx  0.46073866907954264
        \SI{10}{m}    &  VRNN            & $\zb$         &   44.6   \\  % 3stwdgas  0.4601299662745743
        \SI{10}{m}    &  SRNN            & $\zb$         &   47.3   \\  % 2bkvv5i0  0.47317594801349017
        \SI{10}{m}    &  CW-VAE          & $\zb^{(1)}$   &   54.9   \\  % 3m4f3f7z
        \SI{10}{m}    &  STCN            & $\zb^{(2)}$   &   \textbf{42.7}   \\  % rkukk5bn  0,4270625976803488

@@

        1 & WaveNet             & $D_c=96$       & \bfseries 10.96/10.89 & \bfseries 10.85/10.76 & \bfseries 11.12/11.01 & \bfseries 11.05/10.85 \\

@@

        64 & WaveNet            & $D_c=96$       & 13.61/13.24 & 13.58/13.21 & 13.61/13.22 & 13.60/13.21 \\  % 2ghkybtd/3iieasdt

@@

At $s=64$, WaveNet and the LSTM yield significantly worse likelihoods than all LVMs separated by $\sim\SI{1}{bit}$.
The CW-VAE outperforms the VRNN and SRNN when configured with a hierarchy of latent variables. 
With a single layer of latent variables, the CW-VAE is inferior to both SRNN and VRNN but notably still better than the LSTM. 
These observations carry to $s=256$, where a multilayered CW-VAE outperforms the LSTM, VRNN and SRNN. 
The STCN yields the best results at both $s=64$ and $s=256$. 
For strides $s>1$, previous work has attributed the inferior performance of autoregressive models without latent variables, such as WaveNet and the LSTM, to the ability of LVMs to model intrastep correlations \parencite{lai_re-examination_2019}.

@@

\section{Phoneme recognition}\label{sec: phoneme recognition}
Although the likelihood is a practical metric for model comparison and selection, a high likelihood does not guarantee that a model has learned useful representations \parencite{huszar_is_2017}. 
For speech data, we would expect models to learn features related to phonetics which would make them useful for tasks such as automatic speech recognition (ASR). 
The Mel spectrogram is a well-established representation of audio designed for speech recognition and is predefined rather than learned. % \parencite{??}. Basic Correlates of the Auditory Stimulus, https://psycnet.apa.org/record/1951-07758-017, modern Mel scale https://ieeexplore.ieee.org/stamp/stamp.jsp?arnumber=1170013&casa_token=UIyQaj3tBRAAAAAA:GtttZedngBkaxVumjnXctc5zS50fptexPMiWhr-7gDrmIVq3JqJPnGCWYedoymb0X5JTGElmHf8&tag=1,
To assess the usefulness of the representations learned by the benchmarked models, we compare them to the highly useful Mel spectrogram on the task of phoneme recognition. 
Phonemes are fundamental units of speech that relate to how parts of words are pronounced (see also appendix~\cref{app: timit phoneme distributions}).
% rather than characters or words themselves (see also appendix~\cref{apsp: timit phoneme distributions}).

\paragraph{Quantitative}
We train an ASR model to recognize phonemes and compare its performance when using input representations obtained from different unsupervised models. 
For WaveNet and the LSTM, we use the hidden state as the representation. For all LVMs, we use the latent variable. For the hierarchical LVMs and WaveNet, we run the experiment using each possible representation and report only the best one here. 
We compare the learned representations to a log Mel spectrogram with 80 filter banks, hop length 64 and window size 128. 
We also compared to using raw PCM in vectors of 64 elements standardized to $[-1,1]$ but found that the ASR did not reliably converge at all which highlights the importance of input representation. 
The ASR model is a three-layered bidirectional LSTM with 256 hidden units. It is trained with the connectionist temporal classification (CTC) loss \parencite{graves_connectionist_2006} which lets it jointly learn to align and classify without using label alignments. 
We pre"=train all unsupervised models at $s=64$ on the full TIMIT training dataset excluding the validation data (3.7h) as in \cref{tab: likelihoods timit} (left). 
We then train the ASR model on all \SI{3.7}{h} as well as \SI{1}{h} and \SI{10}{m} subsets. We report results on the test set in terms of phoneme error rates (PER) in \cref{tab: phoneme recognition (PER)} (right).
